Tree establishment in models of forest dynamics: a key priority for further research

### 2023-03-13 manuscript/manuscriptRW.Rmd
I remove the right answer wrong reason sentend in discussion

Overall, models are needed to (1) synthesize existing empirical data and
explore their consequences, (2) assess future tree establishment e.g.,
in the context of global change scenarios and, (3) identify most
important processes needed to define relevant ecological processes.
Given the current strategies that are used in models of forest dynamics
to represent tree establishment, model behavior often faces problems
that can have strong impacts on simulated forest dynamics, such as very
high levels of tree establishment, which then necessitate excess
mortality at early stages of tree life to simulate correct stand
structure and composition. Also, inadequately high species diversity in
tree establishment may be simulated. This is characteristic of
"classical" forest gap models [@gutiérrez2016], at least as long as the
simulations comprise a multitude of species. Other models use
calibration against local canopy-level data to define appropriate
ingrowth levels, which is likely to strongly hamper the general
applicability of these models e.g., under scenarios of climate change.
Furthermore, correctly capturing the species composition in tree
establishment as a function of the presence of seed trees in the canopy
is often a particular challenge, potentially leading to unrealistic
successional drift in the model, which must be corrected by factors that
are hard to parameterize [e.g., @lischke2006]. A related issue is the
excessive reduction of species richness due to positive feedback effects
such that eventually just single-species stands remain. This is
sometimes corrected by the incorporation of a low level of seed influx
of all species at all times [@schumacher2006] or by restricting the
number of seeds per species in the seedbank [@lischke2006]. However,
simulated species composition is usually exceedingly sensitive to that
level, whereas the parameters of such functions are poorly constrained
by field data. Lastly, there is often a problem with insufficient
observational constraints on parameter values for models that start from
very small tree sizes (e.g., 10 cm of height or even seeds) and track
tree development in a process-oriented manner by considering a multitude
of ecological influences, rather than emphasizing tree establishment
into a larger size class [e.g., @wehrli2006].

@@

The overarching goal of the experiments was to assess the regeneration
niche as it arises from empirical data against its representation in a
wide range of models of forest dynamics. We define the regeneration
niche as the passing of a breast height diameter threshold of 7 or 10
cm, respectively (synonym: ingrowth). To this end, each modeling group
was provided with a detailed protocol (cf. Supplementary Material 1,
henceforth "SM") with instructions how to perform the simulations,
providing input variables on climate and soil conditions, and the
expected output variables. Neither were further site information (except
for the data specified below) nor any data on tree ingrowth provided
prior to the simulation. That is, the models were run in "blind flight"
mode.

@@

From a modeling point of view, excess ingrowth inevitably requires
excess mortality rates in a later stage, either -- as observed for some
models -- between the 7 and 10 cm diameter thresholds, or soon after the
10 cm threshold has been crossed. In any case, correcting at early
stages for the expected forest densities at later stages is equivalent
to compensating for a first error (excess ingrowth) by a second "error"
(excess mortality). Thus, many models appear to get the 'right answer',
i.e., correct stand structure and composition, for the wrong reasons. It
is highly likely that biased projections will result, because the two
errors are unlikely to be perfectly linked and thus will not always
compensate each other. Hence, this structural problem of most models
investigated here is problematic particularly if the models are to be
used under novel conditions such as under climate change [e.g.,
@huber2021] or in a decision support context [e.g., @thrippleton2021].

@@

The EuFoRIa data as used in this study are unique and well suited to
better understand tree establishment. However, three aspects of these
data may also represent considerable limitations. First, we made a
comparison of tree ingrowth in a equilibrium state, but we cannot assess
how close the forests included in the EuFoRIa dataset were to such a
state. The data were collected in forest reserves where no management
has taken place for long periods of time. This makes our assumption of
an equilibrium between forest properties and environmental drivers more
reasonable than it might appear at first sight. However, the equilibrium
assumption may explain why simulated ingrowth rates were higher than the
observed ones. Second, the data were collected for ingrowth above a 7 cm
threshold, thus limiting the assessment of tree establishment to a
holistic viewpoint. This constituted a hard limit based on which we can
understand some aspects of tree establishment, which in its entirety
often comprises a rather long period since seed production [@price2001].
However, many environmental constraints are acting on young trees
[@käber2021] that we were unable to assess. Yet, ingrowth data with
lower thresholds are simply not available in a harmonized manner across
large environmental gradients. Third, the empirical data were collected
from rather small plots, while we sampled simulated ingrowth rates from
1 ha areas, which may lead to an incorrect representation of space. Even
though this strategy is not ideal, it represents a common challenge when
harmonizing diverse data sources originating from varied sampling
strategies [@portier2022]. It would have been extremely challenging for
such a variety of models to follow a protocol where the spatial sampling
size was different at each of the 200 sites and it would have introduce
additional uncertainty into the results.

>>>>

Overall, models are needed to (1) synthesize existing empirical data and
explore their consequences, (2) assess future tree establishment e.g.,
in the context of global change scenarios and, (3) identify most
important processes needed to define relevant ecological processes.
Given the current strategies that are used in models of forest dynamics
to represent tree establishment, model behavior often faces problems
that can have strong impacts on simulated forest dynamics, such as very
high levels of tree establishment, which then necessitate excess
mortality at early stages of tree life to simulate correct stand
structure and composition. Also, inadequately high species diversity in
tree establishment may be simulated. This is characteristic of
"classical" forest gap models [@gutiérrez2016], at least as long as the
simulations comprise a multitude of species. Other models use
calibration against local canopy-level data to define appropriate
ingrowth levels, which is likely to strongly hamper the general
applicability of these models e.g., under scenarios of climate change.
Furthermore, correctly capturing the species composition in tree
establishment as a function of the presence of seed trees in the canopy
is often a particular challenge, potentially leading to unrealistic
successional drift in the model, which must be corrected by factors that
are hard to parameterize [e.g., @lischke2006]. A related issue is the
excessive reduction of species diversity due to positive feedback
effects such that eventually just single-species stands remain. This is
sometimes corrected by the incorporation of a low level of seed influx
of all species at all times [@schumacher2006] or by restricting the
number of seeds per species in the seedbank [@lischke2006]. However,
simulated species composition is usually exceedingly sensitive to that
level, whereas the parameters of such functions are poorly constrained
by field data. Lastly, there is often a problem with insufficient
observational constraints on parameter values for models that start from
very small tree sizes (e.g., 10 cm of height or even seeds) and track
tree development in a process-oriented manner by considering a multitude
of ecological influences, rather than emphasizing tree establishment
into a larger size class [e.g., @wehrli2006].

@@

The overarching goal of the experiments was to assess the tree
establishment as it arises from empirical data against its
representation in a wide range of models of forest dynamics. We define
the tree establishment as the passing of a breast height diameter
threshold of 7 or 10 cm, respectively (synonym: ingrowth). To this end,
each modeling group was provided with a detailed protocol (cf.
Supplementary Material 1, henceforth "SM") with instructions how to
perform the simulations, providing input variables on climate and soil
conditions, and the expected output variables. Neither were further site
information (except for the data specified below) nor any data on tree
ingrowth provided prior to the simulation. That is, the models were run
in "blind flight" mode.

@@

From a modeling point of view, excess ingrowth inevitably requires
excess mortality rates in a later stage, either -- as observed for some
models -- between the 7 and 10 cm diameter thresholds, or soon after the
10 cm threshold has been crossed. In any case, correcting at early
stages for the expected forest densities at later stages is equivalent
to compensating for a first error (excess ingrowth) by a second "error"
(excess mortality). It is highly likely that biased projections will
result, because the two errors are unlikely to be perfectly linked and
thus will not always compensate each other. Hence, this structural
problem of most models investigated here is problematic particularly if
the models are to be used under novel conditions such as under climate
change [e.g., @huber2021] or in a decision support context [e.g.,
@thrippleton2021].

@@

The EuFoRIa data as used in this study are unique and well suited to
better understand tree establishment. However, three aspects of these
data may also represent considerable limitations. First, we made a
comparison of tree ingrowth in a equilibrium state, but we cannot assess
how close the forests included in the EuFoRIa dataset were to such a
state. The data were collected in forest reserves where no management
has taken place for long periods of time. This makes our assumption of
an equilibrium between forest properties and environmental drivers more
reasonable than it might appear at first sight. However, the equilibrium
assumption may explain why simulated ingrowth rates were higher than the
observed ones. Second, the data were collected for ingrowth above a 7 cm
threshold, thus limiting the assessment of tree establishment to a
holistic viewpoint. This constituted a hard limit based on which we can
understand some aspects of tree establishment, which in its entirety
often comprises a rather long period since seed production [@price2001].
However, many environmental constraints are acting on young trees
[@käber2021] that we were unable to assess. Yet, ingrowth data with
lower thresholds are simply not available in a harmonized manner across
large environmental gradients. Third, the empirical data were collected
from rather small plots, while we sampled simulated ingrowth rates from
1 ha areas, which may lead to an incorrect representation of space. Even
though this strategy is not ideal, it represents a common challenge when
harmonizing diverse data sources originating from varied sampling
strategies [@portier2022]. It would have been extremely challenging for
such a variety of models to follow a protocol where the spatial sampling
size was different at each of the 200 sites and it would have introduce
additional uncertainty into the results. We can not forget here that the
observed data is just a snapshot of an extremely complex and stochastic
reality, which might be the explanation of e.g. lack of clear patterns
when confronted with gradients of climater water balance and degree
days, while the models outputs are the realizations of the definitions
of the regeneration niche.
